Short paper : Automatic predicate testing in formal certification

### 2016-04-27 ProofsAndTests/paper-TAP/content/3-testingInside.tex
paper

However, if the predicate $isSorted$ has been incorrectly written, then the result of this test might inform us that there's something wrong with the formal specification. For example, if we've forgotten the third constructor $consSorted$ in the definition of $isSorted$, then the result of (testSorted 8 4) will be : [False, False, False, False, False, False, False, False] because no produced list will be seen as sorted according to the predicate, and we will therefore know that the predicate $isSorted$ does not capture our intuitive notion of sortedness.

When we want to test $isSorted$ on a big number of tests, we might not want to inspect manually the result of each test. We can write a function $testSorted\_result : (m:Nat)\ \rightarrow\ (n:Nat)\ \rightarrow\ Bool$ that calls $testSorted$ and does the boolean $And$ on each element of the resulting vector. Now, we can for example test the predicate on the first 50 sorted lists of size 9 by running $testSorted\_result\ 50\ 9$ and if we do so we get the overall result $True$ which means that the predicate agrees with the generator on all these 50 tests.

>>>>

When we want to test $isSorted$ on a big number of tests, we might not want to inspect manually the result of each test. We can write a function $testSorted\_result : (m:Nat)\ \rightarrow\ (n:Nat)\ \rightarrow\ Bool$ that calls $testSorted\ m\ n$ and does the boolean $And$ on each element of the resulting vector. Now, we can for example test the predicate on the first 50 sorted lists of size 9 by running $testSorted\_result\ 50\ 9$ and if we do so we get the overall result $True$ which means that the predicate agrees with the generator on all these 50 tests.

However, if the predicate $isSorted$ has been incorrectly written, then the result of this test might inform us that there's something wrong with the formal specification. For example, if we've forgotten the third constructor $consSorted$ in the definition of $isSorted$, then the result of $(testSorted\ 8\ 4)$ will be $False$, which means that at least one of the produced list is not seen as sorted according to $isSorted$, and we will therefore know that this predicate  does not capture our intuitive notion of sortedness.

### 2016-04-27 ProofsAndTests/paper-TAP/content/5-conclusion.tex
paper

We've presented a new way to test a predicate based on an automatic generation of terms that should have the desired property. This adequacy between the predicate and the generator helps to gain confidence in the predicate. The technique presented on section \ref{sect:testingInside} was based on a finite observation of the infinitely many generated terms processed by a decision procedure. We've also shown on an example how these tests can be completely automated. In section \ref{sect:aStepForward} we've presented ideas for going a step forward and replacing these tests on the predicate by some new proofs. These new proofs have the advantage of not requiring the predicate to be decidable. However, they imply that we have to trust these new specifications, which are this time about the predicate itself.

We haven't been able to find much work done in the direction of predicate testing in the environment of proof assistants, but we strongly believe that this aspect is crucial, as there is absolutely no point to prove the correctness of a function relatively to a bad specification. A natural extension of the work presented here is to make it more generic, and to study how proof assistants could provide some primitive bricks that could help to gain confidence in logical properties. A possibility is to provide many robust and generic concepts, like being sorted, so that the user won't have to define his own properties.

>>>>

We've presented a new way to test a predicate based on an automatic generation of terms that should have the desired property. This adequacy between the predicate and the generator helps to gain confidence in the predicate. The technique presented on section \ref{sect:testingInside} was based on a finite observation of the infinite quantity of generated terms processed by a decision procedure. We've also shown on an example how these tests can be completely automated. In section \ref{sect:aStepForward} we've presented ideas for going a step forward and replacing these tests on the predicate by some new proofs. These new proofs have the advantage of not requiring the predicate to be decidable. However, they imply that we have to trust these new specifications, which are this time about the predicate itself.

We haven't been able to find much work done in the direction of predicate testing in the environment of proof assistants, but we strongly believe that this aspect is crucial, as there is absolutely no point to prove the correctness of a function relatively to a bad specification. One could however question why formal certification is needed at all, if after going through all the effort of interactive theorem proving we still have to test the specification itself and need tools to support it. We believe that the process of formalisation precisely helps to uncover things that were missing or weakly specified in the informal requirements. For example, the development of the formally verified compiler CompCert~\cite{Krebbers2014} has contributed to brought to light many under-specified behaviours in the specification of the C standard.

A natural extension of the work presented here is to study how proof assistants themselves could help to gain confidence about predicates and logical formulae. One possible way could be to see how proof assistants could be equipped with an execution engine for formal specifications, that would replace the very specific machinery that we've showed in the running example of this paper. Another possibility, which in fact complements the first, is to have the proof assistant providing many robust and generic concepts once and for all (like the predicate being sorted), because that would save the user from the error prone activity of writing many primitive logical properties. Equipping proof assistants with many generic and useful concepts already available, like bricks ready to be assembled, is one of the current challenge to make proofs assistants really usable.
